Contracts for unions and intersections are hard for Nickel

### 2021-04-13 paper.tex
Typos + TODOs

However, static text-based configuration alone is falling short of expressivity,
and is seldom sufficient. Data cannot be transformed, combined nor shared,
resulting at best in boilerplate and duplication of information, or at worst, in
data being invalid or inconsistent. Data validation isn’t supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
done. Correctly configuring a modern system is hard and failures may have
substantial negative consequences.

This is why the authors have been developping the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while still providing features for writing correct
code and generating valid data. The focus is first and foremost practical:
Nickel originated as a rethink of the language of the Nix package manager, used
to describe one of the biggest sets of software packages\cite{repology}.  Nickel
targets build systems and cloud deployment tools (Kubernetes, Terraform or
NixOps) as well. While we are keen on incorporating existing or doing new
research when it solves a well-defined problem, the general design goals are
always guided by the practical industrial use cases.

@@

Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
Unlike JSON, though, Nickel anticipates large configurations by being both
programmable and typed. In the configuration setting, there is a singular
dilemma with respect to typing: since a program is run on fixed inputs and is
expected to terminate, any relevant type error will show up at evaluation.  Why
bother with the complexity of a static type system? On the other hand, more and
more software systems offload complexity to configurations, in particular with
the infrastructure as code paradigm prevailing today in cloud deployment. When
the complexity of a codebase grows, static types becomes attractive again.  For
reusable code — that is, library functions —, static types are specifically
adapated, and bring all the usual benefits of early error detection, code
robustness, better code structure, documentation, and so on.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Contracts for various extensions of the simply typed $\lambda$-caluclus
have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
varying complexity.  Amongst all the common extensions to the polymorphic lambda
calculus, one of the most simple and ubiquitous turns out to be one of the most
involved: \emph{union types}.

@@

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. However, Nickel is a configuration
language, whose contracts and types system is intended to be expressive enough
to encode data schemas encoutered in practice. Since these schemas do use
untagged unions, there is a strong incentive to support untagged uniohns at the
type level, or at least, at the contract level.

@@

>>>>

However, static text-based configuration alone is falling short of expressivity,
and is seldom sufficient. Data cannot be transformed, combined nor shared,
resulting at best in boilerplate and duplication of information, or at worst, in
data being invalid or inconsistent. Data validation is not supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
done. Correctly configuring a modern system is hard and failures may have
substantial negative consequences.

This is why the authors have been developing the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while still providing features for writing correct
code and generating valid data. The focus is first and foremost practical:
Nickel originated as a rethink of the language of the Nix package manager, used
to describe one of the biggest sets of software packages\cite{repology}.  Nickel
targets build systems and cloud deployment tools (Kubernetes, Terraform or
NixOps) as well. While we are keen on incorporating existing or doing new
research when it solves a well-defined problem, the general design goals are
always guided by the practical industrial use cases.

@@

Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
Unlike JSON, though, Nickel anticipates large configurations by being both
programmable and typed. In the configuration setting, there is a singular
dilemma with respect to typing: since a program is run on fixed inputs and is
expected to terminate, any relevant type error will show up at evaluation.  Why
bother with the complexity of a static type system? On the other hand, more and
more software systems offload complexity to configurations, in particular with
the infrastructure as code paradigm prevailing today in cloud deployment. When
the complexity of a codebase grows, static types becomes attractive again.  For
reusable code — that is, library functions —, static types are specifically
adapted, and bring all the usual benefits of early error detection, code
robustness, better code structure, documentation, and so on.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Contracts for various extensions of the simply typed $\lambda$-calculus
have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
varying complexity.  Amongst all the common extensions to the polymorphic lambda
calculus, one of the most simple and ubiquitous turns out to be one of the most
involved: \emph{union types}.

@@

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. However, Nickel is a configuration
language, whose contracts and types system is intended to be expressive enough
to encode data schemas encountered in practice. Since these schemas do use
untagged unions, there is a strong incentive to support untagged unions at the
type level, or at least, at the contract level.

@@

$(A -> B) /\ (C -> D) ~ (A \/ C) -> (B \/ D)$
$(A \/ B) -> C ~ (A -> C) /\ (B -> C)$

% TODO: lazyness + parallelisation
